Data Specialization for a Staged Lambda Calculus

### 2015-07-05 paper/figures/splitting-overview.tex
filling out explanations

\begin{theorem} \leavevmode
\label{thm:total} 
\begin{itemize} 
\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
\item For term $e$, if $\typesone e A$, then $\splitone e A c {l.r}$.
\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
\end{itemize}
\end{theorem}

\begin{theorem} \leavevmode
\label{thm:unique}
\begin{itemize} 
\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
\item If $\splitone e A p {l.r}$ and $\splitone e A {c'} {l'.r'}$, then $c = c'$, and $l.r = l'.r'$.
\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
\end{itemize}
\end{theorem}

\begin{theorem}
\label{thm:allCorrect}
If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
then $\reduce p b$ and $[b/l]r \equiv q$.
\end{theorem}

\begin{lemma}
\label{lem:twoCorrect}
If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}

\begin{lemma}
\label{lem:oneCorrect}
If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}

\begin{lemma}
\label{lem:oneCorrect}
If $e \stepsym \bbonep e'$ and $\reduce {e'} i$, then $\reduce e i$.
\end{lemma}

@@

\end{abstrsyn}
\caption{\TODO put a real caption here and format the figure nicely}
\label{fig:splittingSummary}
\end{figure*}

>>>>

\begin{theorem} [Splitting Totality] \leavevmode
\label{thm:total} 
\begin{itemize} 
\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
\item For term $e$, if $\typesone e A$, then $\splitone e A c {l.r}$.
\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
\end{itemize}
\end{theorem}

\begin{theorem} [Splitting Uniqueness] \leavevmode
\label{thm:unique}
\begin{itemize} 
\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
\item If $\splitone e A p {l.r}$ and $\splitone e A {c'} {l'.r'}$, then $c = c'$, and $l.r = l'.r'$.
\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
\end{itemize}
\end{theorem}

\begin{theorem} [Program Correctness]
\label{thm:allCorrect}
If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
then $\reduce p b$ and $[b/l]r \equiv q$.
\end{theorem}

\begin{lemma} [\bbtwo\ Step Correctness]
\label{lem:twoCorrect}
If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}

\begin{lemma} [\bbonem\ Step Correctness]
\label{lem:oneCorrect}
If $e \stepsym \bbonem e'$, $\splitone e A c {l.r}$, $\splitone {e'} A {c'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}

\begin{lemma} [Hoisting Step Correctness]
\label{lem:hoistCorrect}
If $\lift e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, then $c = c'$, and $l.r \equiv l'.r'$.
\end{lemma}

@@

\end{abstrsyn}
\caption{Splitting theorems.}
\label{fig:splittingSummary}
\end{figure*}

### 2015-07-05 paper/semantics.tex
filling out explanations

\input{figures/dynamics}

@@

\TODO pull out the theorems into a figure

The statics and dynamics of \lang\ are related by a type safety theorem
comprising progress and preservation, again annotated by world.
In all cases below, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
\begin{theorem}\leavevmode
\label{thm:progress} 
\begin{itemize} 
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
\end{theorem}
\begin{theorem}\leavevmode
\label{thm:preservation} 
\begin{itemize} 
\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}
\end{theorem}

>>>>

\input{figures/dynamics}
\input{figures/dynamics-overview}

@@

The statics and dynamics of \lang\ are related by a type safety theorems
in figure \ref{fig:typeSafety}, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.

### 2015-07-05 paper/splitting.tex
filling out explanations

% Splitting statically transforms terms in \lang\ into a pair of
% monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
% and second-stage computations in the original term.

% Recall that full evaluation (both the first and second pass) of a term $\coltwo e A$ 
% only produces its ``final result'' at the second stage.
% In contrast, \bbonem\ terms may have final results at both stages. 
% For example, consider the term:
% \begin{lstlisting}
% 1`(`3`gr{`1`4+5`3`}`1`, next{`2`6+7`1`})`
% \end{lstlisting}

% Thus, the idea of splitting for \bbtwo\ terms
% is to generate terms where evaluating $f_1$ in a first pass, and then
% evaluating $f_2$ on the result of the first pass, yields the same
% value as full evaluation of the original \lang\ program.

% first pass evaluation of
% $f_1$ should produce the same value as first-stage evaluation of the
% original \lang\ term.  However, complexity arises because
% \bbonem\ terms Thus we also want
% the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
% the same value as fully evaluating all second-stage computations in
% the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
% which only produce values in the first stage, we only need to compare
% the result of $f_1$ with that of the \lang\ term.

% Handling the asymmetry between the results of evaluating
% \bbonem\ terms (which may produce results at both the first and second
% stage) and \bbtwo\ terms (which only produce second-stage results) is
% fundamental to the structure of the splitting algorithm presented in
% this section.

Our splitting algorithm is given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$,
which apply to \bbtwo\ terms, \bbonem\ terms, and partial values, respectively.
Before discussing the meaning of these outputs, 
we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting relation is a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
The first part of the masked value, $i$, is a monostage value holding
the result of all first-stage components of $v$.  The second
component, $q$, is a residual representing all second-stage
variable references of $v$.

To provide intuition about the intent of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`yhat`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:valMask}.

@@

\TODO why the added complexity for \bbonem\ terms?

For any $\colmix e A$, 
splitting $e$ (via $\splitonesym$) again 
yields the pair of monostage terms $\pipeM c l r$. 
Evaluating $c$ in the first pass produces the pair
$\tup{i,b}$, where the first element $i$ represents the first stage part of the results of $e$,
and $b$ is the boundary value.
Since execution of the first pass both serves to generate input for
the second pass, we say $c$ is a {\em combined term} that produces a
\emph{immediate result} ($i$) and a boundary value ($b$) for the resumer ($l.r$).

The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ is given in \ref{lem:oneCorrect}.
Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
But additionally, it also requires that first stages results be exactly identical.

\TODO talk about lifting step correctness

\TODO talk about ground case

@@

% For example, in the simple case where $e$
% contains entirely first-stage computation, such as:
% \begin{lstlisting}
% 1` 3`gr{`1`literalone+literaltwo`3`}`
% \end{lstlisting}
% the first-stage result is the value 3.  The second element of the pair, $b$,
% is the boundary value used as input to the resumer in second pass
% evaluation as $[b/l]r \equiv q$, where $q$ is the same as
% result produced by second-stage computations in $e$.  For example, in
% the case where $\rtab \xi v$ contains entirely second-stage
% computation, such as:
% \begin{lstlisting}
% 2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
% \end{lstlisting}
% the second-stage result is the value 5. Since in general residual
% tables and partial values describe a mixture of both first and second
% stage computations, in the subsequent subsection we define an
% operation, called {\em masking}, that separates residual tables and
% partial values into their first- and second-stage results. We then
% describe the algorithm for splitting both \bbonem\ terms
% ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

% %Since $c$ ensapulates computations that produce both $e$'s first-stage
% %result and also the boundary value required by the resumer, we refer
% %to it as a {\em combined term}.

% %Note that since $\coltwo e A$ reduces to an entirely second-stage
% %residual, its first-stage computations only exist to generate input
% %for second-stage computations.  In contrast, since $\colmix e A$
% %reduces to a multi-stage residual table and partial value, its first
% %stage computations serve to both generate the value used as input for
% %the second stage {\em and also} produce a first-stage value.
% %Correspondingly, the precomputation $p$ resulting from splitting
% %\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
% %the resumer; but the combined term $c$ resulting from splitting
% %\bbonem\ terms reduces to a tuple containing both the first stage {\em
% %  immediate result} $i$ and the boundary value $b$ (motivating the
% %name {\em combined term}).

>>>>

The goal of a stage splitting translation is to send programs in a multistage language
to a form where the stages are seperated at the top level.
To be valid, such a translation must meet two requirements:
at values it must not lose information,
and at terms it must respect the semantics.

For a two-stage language like \lang, this means translating a multistage
object into a pair of monostage objects
which respectively embody the first-stage and second-stage content of the original.
Since \lang\ has three classes of multistage object, 
our splitting algorithm has three forms: 
\bbtwo-term splitting, \bbonem-term splitting, and partial value splitting.
These are given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$.
The rules of all three judgements are given in \ref{fig:termSplit}.

Before discussing the validity of these translations, 
we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting relation is a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
The first component of the split value, the \emph{immediate value} $i$, is a monostage value holding
the result of all first-stage components of $v$.  The second
component, $q$, is a residual representing all second-stage
variable references of $v$.

To provide intuition about the intent of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`yhat`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:valMask}.
Observe that partial value splitting does not lose any meaningful information (namely injection tags),
which is the validity requirement for a translation of value forms.

@@

For any $\colmix e A$, 
splitting $e$ (via $\splitonesym$) again 
yields the pair of monostage terms $\pipeM c l r$.

The responsibilities of this first stage component, $c$, can be seen as a combination of the
those of world \bbtwo\ term splitting and partial value splitting.
Since $e$ is a term, the first stage computations within it may yet still
affect its second stage computations,
and so $c$ needs to produces a boundary value $b$ to be passed to the resumer ($l.r$).
But since $e$ is a world \bbonem\ construct, it has an eventual result at the 
first stage as well as the second, and so it must produce an immediate value $i$.
The term $c$ meets both of these responsibilities by reducing to the tuple $\tup{i,b}$,
and so we call it a {\em combined term}.

The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.
Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
But additionally, it also requires that first stage results be exactly identical.

The correctness of \bbonem\ term splitting with respect to hoisting steps is given in \ref{lem:hoistCorrect}.
Because hoisting steps are nothing but rearrangement of second stage code,
this lemma can use the strong requirement of idnetical combined terms.

@@

\paragraph{\bbtwo\ Compatility.}
\paragraph{\bbone\ Compatility.}
\paragraph{Functions.}
\paragraph{Next and prev.}
\paragraph{Ground and Caseg.}
